---
title: "Case_Study1"
author: "Emre Can Çamcı"
date: "`r Sys.Date()`"
output: 
  html_document : default
  pdf_document : default
---

## Ask Phase

-   Business task is to understand how annual members and casual riders use Cyclistic bikes differently

## Prepare Phase

-   To analyze the yearly trips from 2022 data has been downloaded from this [link](https://divvy-tripdata.s3.amazonaws.com/index.html) and has been made available by Motivate International Inc. under this [licence](https://ride.divvybikes.com/data-license-agreement).

## Process Phase

Data will contain millions of rows and to process faster I used R.

#### Importing data to R

-   Loading libraries

```{r libraries, warning=FALSE, include=FALSE}
library(tidyverse) # For data import & wrangling
library(lubridate) # For date functions
library(scales)    # For better visualizations

```

-   Importing the cyclist data of 2022

```{r monthly data, warning=FALSE}
jan_2022 <- read.csv("202201-divvy-tripdata.csv")
feb_2022 <- read.csv("202202-divvy-tripdata.csv")
mar_2022 <- read.csv("202203-divvy-tripdata.csv")
apr_2022 <- read.csv("202204-divvy-tripdata.csv")
may_2022 <- read.csv("202205-divvy-tripdata.csv")
jun_2022 <- read.csv("202206-divvy-tripdata.csv")
jul_2022 <- read.csv("202207-divvy-tripdata.csv")
aug_2022 <- read.csv("202208-divvy-tripdata.csv")
sep_2022 <- read.csv("202209-divvy-publictripdata.csv")
oct_2022 <- read.csv("202210-divvy-tripdata.csv")
nov_2022 <- read.csv("202211-divvy-tripdata.csv")
dec_2022 <- read.csv("202212-divvy-tripdata.csv")

```

-   Comparing column names of each data to check if they match perfectly before joining them into a data frame.

```{r column name check, warning=FALSE, include=FALSE}
colnames(jan_2022)
colnames(feb_2022)
colnames(mar_2022)
colnames(apr_2022)
colnames(may_2022)
colnames(jun_2022)
colnames(jul_2022)
colnames(aug_2022)
colnames(sep_2022)
colnames(oct_2022)
colnames(nov_2022)
colnames(dec_2022)

```

-   Creating the data frame with the monthly data.

```{r dataframe, warning=FALSE, include=FALSE}
yearly_trips <- bind_rows(jan_2022,feb_2022,mar_2022,apr_2022,may_2022,jun_2022,
                        jul_2022,aug_2022,sep_2022,oct_2022, nov_2022,dec_2022)

```

-   Dropping columns which will not be used

```{r column drop, warning=FALSE, include=FALSE}
yearly_trips <- yearly_trips %>% 
  select(-c(start_lat,start_lng,end_lat,end_lng))
```

-   Checking the new data frame.

```{r dataframe check, warning=FALSE, include=FALSE}
colnames(yearly_trips) # List of column names
nrow(yearly_trips) # Row number in data frame
head(yearly_trips, 10) # To see the first 10 row of data frame
tail(yearly_trips, 10) # To see the last 10 row of data frame
str(yearly_trips) # List of columns and data types
summary(yearly_trips) #Statistical summary of data

```

## Analyze Phase

-   To further analyze the data adding date, weekday, month, and a ride length column.

```{r column add}
#Adding a ride length column to analyze the duration of the rides in seconds
yearly_trips$ride_length <- difftime(yearly_trips$ended_at,yearly_trips$started_at)
yearly_trips$date <- as.Date(yearly_trips$started_at)
yearly_trips$month <- format(as.Date(yearly_trips$date), "%m")
yearly_trips$day_of_week <- format(as.Date(yearly_trips$date), "%A")
```

-   Checking the created column

```{r ride_length check}
str(yearly_trips)
head(yearly_trips, 10)
tail(yearly_trips, 10)
summary(yearly_trips)
```

-   Converting c(ride_length) to numeric so calculations can be executed.

```{r ride_lenght format}
yearly_trips$ride_length <- as.numeric(as.character(yearly_trips$ride_length))
is.numeric(yearly_trips$ride_length) # To check the class of column
summary(yearly_trips)
```

-   After checking summary, c(ride_length) has some negative values and has to be excluded.

```{r remove negative values}
yearly_trips_v2 <- subset(yearly_trips, yearly_trips$ride_length >0)
summary(yearly_trips_v2)
```

#### Organizing Data

-   Reordering and dropping the columns to understand data easier

```{r organizing}
#Removing columns which will not be use to analyze
yearly_trips <- yearly_trips %>% 
  select(-c(started_at,ended_at,start_station_name,start_station_id,
            end_station_name,end_station_id))
#Reordering columns
colnames(yearly_trips)
col_order <- c("ride_id", "rideable_type","date", "day_of_week","month","ride_length","member_casual")
yearly_trips_v3 <- yearly_trips_v2[, col_order]
```

-   Converting c(ride_length) to minutes

```{r sec to min}
yearly_trips_v3$ride_length <- yearly_trips_v3$ride_length /60
```

#### Analyzing the data

To understand how the member and casuals use bikes differently checking:

* Ride duration
* Number of rides


##### Ride duration

* First a descriptive analysis on ride duration.

```{r des analysis}
yearly_trips_v3 %>% 
  summarise(min(ride_length),mean(ride_length),max(ride_length))
```

We can see that average ride length is 19.4 minutes per rider.

* Comparing members and casual users monthly average ride duration.

```{r monthly avg duration, message=FALSE}
yearly_trips_v3 %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n(), average_duration=mean(ride_length)) %>% 
  arrange(member_casual,month) %>% 
  ggplot(aes(x=month, y=average_duration, fill= member_casual))+
  geom_col(position = "dodge")+
  geom_hline(yintercept = 19.4)+
  xlab("Months")+
  scale_x_discrete(labels = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug",
                              "Sep","Oct","Nov","Dec"))+
  ylab("Average duration(min)")+
  labs(title = "Monthly average ride duration in minutes" , fill ="Rider Type")
```

The plot shows that casuals tends to ride longer than members, and members ride duration is below average ride duration.

* If we look at weekly average ride duration

```{r weekly avg duration, message=FALSE}
yearly_trips_v3 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n(), average_duration=mean(ride_length)) %>% 
  arrange(member_casual,day_of_week) %>% 
  ggplot(aes(x=day_of_week, y=average_duration, fill= member_casual))+
  geom_col(position = "dodge")+
  geom_hline(yintercept = 19.4)+
  xlab("Weekday")+
  ylab("Average duration(min)")+
  labs(title = "Weekly average ride duration in minutes" , fill ="Rider Type")
```

Days aren't ordered to fix it

```{r day order}
yearly_trips_v3$day_of_week <- ordered(yearly_trips_v3$day_of_week, levels= 
                                         c("Monday","Tuesday", "Wednesday", 
                                           "Thursday", "Friday", "Saturday",
                                           "Sunday"))
```

After fixing it

```{r weekly avg duration2, message=FALSE}
yearly_trips_v3 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n(), average_duration=mean(ride_length)) %>% 
  arrange(member_casual,day_of_week) %>% 
  ggplot(aes(x=day_of_week, y=average_duration, fill= member_casual))+
  geom_col(position = "dodge")+
  geom_hline(yintercept = 19.4)+
  xlab("Weekday")+
  ylab("Average duration(min)")+
  labs(title = "Weekly average ride duration in minutes" , fill ="Rider Type")
```

##### Number of rides

* Lets look at yearly trips.

```{r yearly trips}
yearly_trips_v3 %>% 
  summarise(number_of_rides = n())

```

This year total of 5.667.186 trips made by our members and casuals.

* If we look at our monthly trips.

```{r monthly rides vis, message=FALSE}
yearly_trips_v3 %>% 
  group_by(month, member_casual) %>% 
  summarise(number_of_rides = n()) %>% 
  ggplot(aes(x=month,y=number_of_rides, fill=member_casual))+
  geom_col(position = "dodge")+
  xlab("Month")+
  scale_x_discrete(labels = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug",
                              "Sep","Oct","Nov","Dec"))+
  ylab("Number of rides")+
  scale_y_continuous(labels = comma)+
  labs(title="Monthly rides by rider type", fill= "Rider Type")

```

From the plot we can see that casuals tends to rent more bike in summer.

* If we look at weekly trips.

```{r weekly rides vis, message=FALSE}
yearly_trips_v3 %>% 
  group_by(day_of_week, member_casual) %>% 
  summarise(rider_count = n()) %>% 
  ggplot(aes(x=day_of_week,y=rider_count, fill=member_casual))+
  geom_col(position = "dodge")+
  xlab("Weekdays")+
  ylab("Number of rides")+
  scale_y_continuous(labels = comma, breaks = seq(0,500000, by =100000))+
  labs(title="Weekly rides by rider type", fill= "Rider Type")
```

In contrast to ride duration our members rent more bike and casual rent more bike at weekends.

## Share Phase

After the analysis:

* Casuals ride longer than members.

* Members rent more bike than casuals.

* Casuals tend to rent more bike in summer and weekends.

## Recommendations

* A discount campaign can be launched for membership in the summer since the casual usage rate increases.

* Discount coupons can be issued according to the length of their trips.
